---
title: "FinalProj"
author: "Cong Yu"
date: "4/19/2018"
output: html_document:
  code_folding: hide

---

* loading and exploying data
* individual feature analysis
    +
    +
* Data Cleaning

* Features relation
    +
    +
* build new features

* external datasets
    + weather impact


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message = FALSE}
library('ggplot2') # visualisation
library('scales') # visualisation
library('grid') # visualisation
library('RColorBrewer') # visualisation
library('corrplot') # visualisation
library('alluvial') # visualisation
library('dplyr') # data manipulation
library('readr') # input/output
library('data.table') # data manipulation
library('tibble') # data wrangling
library('tidyr') # data wrangling
library('stringr') # string manipulation
library('forcats') # factor manipulation
library('lubridate') # date and time
library('geosphere') # geospatial locations
library('leaflet') # maps
library('leaflet.extras') # maps
library('maps') # maps
library('xgboost') # modelling
library('caret') # modelling
library(ggthemes)
library(knitr)
library(scales)
library(ggmap)
library(RColorBrewer)
library(treemap)
# gis packages
library(sp)
library(rgeos)
library(geosphere)

```
## Data loading
```{r warning=FALSE, results=FALSE}
train <- as.tibble(fread('train.csv'))
test <- as.tibble(fread('test.csv'))
full <- bind_rows(train %>% mutate(dset = "train"), 
                     test %>% mutate(dset = "test",
                                     dropoff_datetime = NA,
                                     trip_duration = NA))
```

## Data Exploying
```{r}
summary(train)
glimpse(train)
glimpse(test)
```


## Features summary

- 'vender id' has values 1 and 2, represents two differernt taxi companies in NYC. We should treat it as a factor.  

- 'pickup datetime' is in both train and test dataset, 'dropoff datetime' only in train dataset.  

-'passenger count', number of passengers, has values from 0 to 9. We should treat it as a factor  

-'pickup longtitude&latitude' are in both train and test dataset, 'dropoff longtitude&latitude' only in train set.  

-'store and fwd flag', 'N' means trip data was sent immediately to server. 'Y' means trip data was held in the memory there was no connection to server.  

-'trip duration' is only in train dataset.  

## individual feature analysis
```{r}
# take a look at trip duration 

# rect <- data.frame(xmin=1e+04, xmax=1e+05, ymin=-Inf, ymax=Inf) 
#fail to put two rectangles on outliers
ggplot(data = train, aes(x = trip_duration))+
  geom_histogram(bins = 100, color = "green")+
  geom_vline(xintercept = 10, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = 3600*22, linetype = "dashed", color = "red", size = 1) + 
  scale_x_log10()+
  scale_y_sqrt()


```

- it's generally normal distributed  

- strange values that should be removed s.t. less than 10 sec trips and longer than 22 hours trips   


```{r}
par(mfrow=c(1,1))
# take a  look at passenger count
ggplot(data = train, mapping = aes(x = passenger_count))+
  geom_bar( fill  = "blue")
#exact counts
table(train$passenger_count)
# mean and median of passenger count
boxplot(train$passenger_count,col ="pink", ylab = "passenger_count", main= "passenger_count,mean(magenta),median(red)")
abline(h = mean(train$passenger_count),col= "magenta")
abline(h = median(train$passenger_count, col = "red", lwd = 2))
```

- the majority of passenger number is 1, some 0 and over 8 passengers clearly not reasonable and should be removed.  


```{r}
# two different taxi companies
ggplot(data = train, mapping =aes(x = vendor_id ) )+
  geom_bar()
# exact counts
table(train$vendor_id)
```

- we can see vendor 2 has more trips than vendor 1, I guess it owns more taxis.  

```{r}
# store and fwd flag
ggplot(data = train, mapping =aes(x = store_and_fwd_flag ) )+
  geom_bar()
# exact numbers
table(train$store_and_fwd_flag)

```
- majority of trips sent data imeediately.

```{r}
#pickup and dropoff datetime
#datetime formate transform
train$pickup_datetime = ymd_hms(train$pickup_datetime)
train$dropoff_datetime = ymd_hms(train$dropoff_datetime)
train$month<- lubridate::month(train$pickup_datetime, label = TRUE) 
train$wday <- lubridate::wday (train$pickup_datetime, label = TRUE)
train$hour <- lubridate::hour(train$pickup_datetime)
train$date <- lubridate::day(train$pickup_datetime)
train$minutes <-lubridate:: minute(train$pickup_datetime)
#pickup
ggplot(data = train, mapping =aes(x = pickup_datetime ) )+
   geom_histogram(fill = "blue", bins = 120)
#dropoff
ggplot(data = train, mapping =aes(x = dropoff_datetime ) )+
geom_histogram(fill = "blue", bins = 120)
# by month
ggplot(data = train, mapping =aes(x = month ) )+
  geom_bar(fill = "green")

# by days of a week
ggplot(data = train, mapping =aes(x = wday ) )+
  geom_bar(fill = "green")
# by date
ggplot(data = train, mapping =aes(x = date ) )+
  geom_bar(fill = "blue")
# by hour
ggplot(data = train, mapping =aes(x = hour ) )+
  geom_bar(fill = "blue")
```
- the pickups and dropoffs are nearly identical since the trips are likely start and end on same day.  

- through the year, March and April have slightly more trips than other months.  

- through the week, Monday to Friday have trips than increased linearly, then Saturday starts to decrease.  

- through the month, the last five days tend to have less trips.  

- through the day, clearly least trips from midnight to 6 A.M. The peaks are around 6 P.M.  

```{r, warning=FALSE}
#long and lat visualization in NYC
min_lat <- 40.6
max_lat <- 40.9
min_long <- -74.05
max_long <- -73.7

ggplot(train, aes(x=pickup_longitude, y=pickup_latitude)) +
geom_jitter(size=0.6, color = "white") +
scale_x_continuous(limits=c(min_long, max_long)) +
scale_y_continuous(limits=c(min_lat, max_lat))+
theme_dark()
# passenger > 5
ggplot(train %>% 
         filter(passenger_count>5), 
         aes(x=pickup_longitude, y=pickup_latitude,  color=vendor_id)) +
         geom_point(size=0.6, color="#999999") +
         stat_summary_hex(fun = sum, bins=100, alpha=0.7) +
         scale_x_continuous(limits=c(min_long, max_long)) +
         scale_y_continuous(limits=c(min_lat, max_lat)) +
         theme_dark() +
         scale_fill_gradient(low="#CCCCCC", high="#27AE60", labels=seconds) +
         labs(title = "Pickup locations for passenger more than 5") +
         coord_equal()
```


- We can see majority trips took place in NYC Center Area. Trips with more than 5 passengers distributed evenly in town.   



```{r}
#check data points that are fake: not in NYC area

#load map data
state <- map_data("state")
nyc <- state %>% filter(region=="new york")
county <- map_data("county")
nyc_county <- county %>% filter(region=="new york")

set <- rbind(
  train %>% dplyr::select(pickup_longitude, pickup_latitude) %>% mutate(set="tr"),
  test %>% dplyr::select(pickup_longitude, pickup_latitude) %>% mutate(set="te"))


#we can check some coords are outside the United States as well as NYC

#filter coords outside NYC 
out <- train %>%
  filter((pickup_longitude > max(nyc$long) | pickup_longitude < min(nyc$long)) | (pickup_latitude > max(nyc$lat) | pickup_latitude < min(nyc$lat)))

#See those coords in details with real map 
leaflet(data=out, width="100%") %>%
  addTiles() %>%
  addMarkers(~pickup_longitude, ~pickup_latitude, popup=as.character(out$id))


```

- Now we get those couple "fake" data points and they will be removed later in data cleaning.  


## feature relation

```{r}
#vender id vs others
#month vs vender id
train %>%
ggplot( aes(x = month, fill = as.factor(vendor_id)))+
  geom_bar(position ="dodge")+
  labs(y = "count", x = "Months")
# days of a week vs vender id
train %>%
ggplot(aes(x = wday, fill = as.factor(vendor_id)))+
geom_bar(position ="dodge")+
labs(y = "Total number of pickups", x = "Days of the week")
# hours of a day vs vender id
train %>%
ggplot(aes(x = hour, fill = as.factor(vendor_id)))+
geom_bar()+
labs(y = "Total number of pickups", x = "Hours")
# trip duation vs vender id
train %>%
  ggplot(aes(trip_duration, fill =  as.factor(vendor_id))) +
  geom_density(position = "stack") +
  scale_x_log10()
```

- each of the vender has trips distribute similarly to overall trips,  
regarding to month, days of a week, hours in a day.   

- vendor 1 has most of it's trips last about 15-17 mins.  

```{r}
#hours of a day vs others
#hours of a day vs month
train %>%
ggplot(aes(x = hour,fill = month))+
  geom_bar(position = "dodge")+
  facet_grid(~month, scales = "free")+
  scale_fill_manual(labels= levels(month), values=c("red","blue","green","brown","pink","magenta"))+ 
  labs(y = "Total number of pickups", x = "Hours")
#hours of a day vs days of a week
train %>%
ggplot(aes(x = hour,color= wday))+
  geom_bar(position = "dodge")+
  facet_grid(~wday, scales = "free")+
  scale_fill_manual(labels= c("sun","mon","tue","wed","Thur","fri","sat"),values=c("red","blue","green","brown","pink","magenta","yellow"))+ 
  labs(y = "Total number of pickups", x = "Hours")
#hours of a day vs minuetes in an hour
train %>%
ggplot(aes(x = minutes,fill=as.factor(hour)))+
  geom_bar(position = "dodge")+
  facet_grid(~ hour, scales = "free")+
  labs(y = "Total number of pickups", x = "Minutes")
```

- In these plots we can see hourly distribution in each day of a week/ each month,  
or hourly distribution detailed in minutes.  

- not much different patterns compare to previous plots.  

- We can find from Monday to Sunday, the peaks are moving from early night to late night.

#build new features
```{r}
#direct distance for a trip
pick_coord <- train %>%
  dplyr::select(pickup_longitude, pickup_latitude)
drop_coord <- train %>%
  dplyr::select(dropoff_longitude, dropoff_latitude)
train <- train %>%
  mutate(dist = distCosine(pick_coord, drop_coord))


train %>%
  ggplot(aes(dist, trip_duration)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Direct distance [m]", y = "Trip duration [s]")

# after excluding extreme
train %>%
  filter(trip_duration < 3600 & trip_duration > 120) %>%
  filter(dist > 100 & dist < 100e3) %>%
  ggplot(aes(dist, trip_duration)) +
  geom_bin2d(bins = c(500,500)) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Direct distance [m]", y = "Trip duration [s]")
```

- It is reasonable that duation of a trip has positive linear relationship with direct distance.  


```{r}
# speed in a trip 
train <- train %>%
  mutate(speed = dist/trip_duration*3.6)
train %>%
  filter(speed > 2 & speed < 1e2) %>%
  ggplot(aes(speed)) +
  geom_histogram(fill = "red", bins = 50) +
  labs(x = "Average speed ")

train %>%
  group_by(wday, vendor_id) %>%
  summarise(median_speed = median(speed)) %>%
  ggplot(aes(wday, median_speed, color = vendor_id)) +
  geom_point(size = 4) +
  labs(x = "Day of the week", y = "Median speed [km/h]")

train %>%
  group_by(hour, vendor_id) %>%
  summarise(median_speed = median(speed)) %>%
  ggplot(aes(hour, median_speed, color = vendor_id)) +
  geom_smooth(method = "loess", span = 1/2) +
  geom_point(size = 4) +
  labs(x = "Hour of the day", y = "Median speed [km/h]") +
  theme(legend.position = "none")

train %>%
  group_by(wday, hour) %>%
  summarise(median_speed = median(speed)) %>%
  ggplot(aes(hour, wday, fill = median_speed)) +
  geom_tile() +
  labs(x = "Hour of the day", y = "Day of the week") +
  scale_fill_distiller(palette = "Spectral")


```

- Since the distance used to calculate speed is direct distance from two points,  
the average speed is clearly under estimated. But it's reasonable for trips in NYC to be relatively slower than other places.  

- It is interesting that taxis go slower from Sunday to Thursday then start to be faster.  

- It's an easy guess that speed will be the fastest around 5 since there will be least traffic in a day.   

- In heatmap each day in a week has its fastest speed around 5 a.m.  


```{r}
#direction
train$bearing = bearing(pick_coord, drop_coord)

train %>%
  filter(dist < 1e5) %>%
  ggplot(aes(bearing, dist)) +
  geom_bin2d(bins = c(100,100)) +
  labs(x = "Bearing", y = "Direct distance") +
  scale_y_log10() +
  theme(legend.position = "none") +
  coord_polar() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45))

train %>%
  filter(trip_duration < 3600*22) %>%
  filter(dist < 1e5) %>%
  ggplot(aes(bearing, trip_duration)) +
  geom_bin2d(bins = c(100,100)) +
  scale_y_log10() +
  labs(x = "Bearing", y = "Trip duration") +
  coord_polar() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45))

train %>%
  filter(speed < 75 & dist < 1e5) %>%
  ggplot(aes(bearing, speed)) +
  geom_bin2d(bins = c(100,100)) +
  labs(x = "Bearing", y = "Speed") +
  coord_polar() +
  scale_x_continuous(breaks = seq(-180, 180, by = 45))



```





#DATA CLEANING
```{r}
#remove coords outside NYC
# wired trips with duration < 10 sec or > 22 hours

train <- train %>%
  filter(!id %in% out$id,
         trip_duration < 22*3600,
         trip_duration > 10,
         dist > 0 | (near(dist, 0) & trip_duration < 60),
         speed < 100)
```

#external datasets
```{r}
weather <- as.tibble(fread("/Users/yucong/Downloads/datawrangling/FinalProject/weather_data_nyc_centralpark_2016(1).csv"))
# Reformating and cleaning weather dataset
weather <- weather %>%
  mutate(date1 = dmy(date),
         rain = as.numeric(ifelse(precipitation == "T", "0.01", precipitation)),
         s_fall = as.numeric(ifelse(`snow fall` == "T", "0.01", `snow fall`)),
         s_depth = as.numeric(ifelse(`snow depth` == "T", "0.01", `snow depth`)),
         all_precip = s_fall + rain,
         has_snow = (s_fall > 0) | (s_depth > 0),
         has_rain = rain > 0,
         max_temp = `maximum temperature`,
         min_temp = `minimum temperature`)
add_Info <- weather %>%
  select(date1, rain, s_fall, all_precip, has_snow, has_rain, s_depth, max_temp, min_temp)
train <- train %>%
  mutate(date1 = date(pickup_datetime))

train <- left_join(train, add_Info, by = "date1")


train %>%
  group_by(date1) %>%
  summarise(trips = n(),
            snow_fall = mean(s_fall),
            rain_fall = mean(rain),
            all_precip = mean(all_precip)) %>%
  ggplot(aes(date1, snow_fall)) +
  geom_line(color = "blue", size = 1.5) +
  labs(x = "", y = "Snowfall") +
  scale_y_sqrt() +
  scale_x_date(limits = ymd(c("2015-12-28", "2016-06-30")))

train %>%
  group_by(date1) %>%
  summarise(trips = n(),
            snow_depth = mean(s_depth)) %>%
  ggplot(aes(date1, snow_depth)) +
  geom_line(color = "purple", size = 1.5) +
  labs(x = "", y = "Snow depth") +
  scale_y_sqrt() +
  scale_x_date(limits = ymd(c("2015-12-29", "2016-06-30")))

ggplot(data = train, mapping =aes(x = pickup_datetime ) )+
   geom_histogram(fill = "blue", bins = 120)

train %>%
  group_by(date1) %>%
  summarise(median_speed = median(speed)) %>%
  ggplot(aes(date1, median_speed)) +
  geom_line(color = "orange", size = 1.5) +
  labs(x = "Date", y = "Median speed")


```


















